\textls[-0]{\thetitle}

### 2023-03-13 metatheory-condensed.tex
Done Changing to named variables.
Remove mention of the universe level in main text
Change Back the Rule Name.

\cref{fig:mltt-selected} presents the syntax and selected typing rules of MLTT %(predicative) MLTT
(and \cref{fig:fmltt-selected} \TT).
Dependent function types \nTyPi{x}{A}{B},
dependent product types \nTySigma{x}{A}{B}, and
identity types \TyId{t_1}{t_2} are standard.
So are their introduction and elimination forms.
Capture-avoiding substitution is notated $\bullet[\bullet/x]$.
A singleton type $\TyS{t}$ helps expose the definition of a term~$t$
in its type (rule \ruleref{n/tm/s})~\cite{aspinall1995singleton,stone2000}.
For each derivation (of each judgment) we introduce judgmental equality
(of the form \goodTerm{\Gm}{t_1 \equiv t_2}{T})
indicating convertibility.

We use Tarski-style universes \cite{hofmann1997syntax},
which, unlike Russell-style ones,
are not inhabited by types directly but rather by the \emph{codes} of types
and arguably behave better \cite{luo2012notes}.
Term $\codety{T}$ encodes type~$T$ (\ruleref{n/tm/c}) and thus inhabits universe~$\cU$.
Type $\El{t}$ decodes term~$t$ (\ruleref{n/ty/el}).
%\footnote{By contrast,  Russel-style universes are inhabited by types directly.  Tarski-style universes are ``semantically more fundamental'' \cite{luo2012notes} and friendlier to our meta-theoretical reasoning.}
There is an infinite hierarchy of universes;
we omit universe levels in the presentation for conciseness.
%\rA{Why is W(t) a term and not a type?}\YZreply{I think the reviewer is wondering what is gained by encoding and decoding types, compared to types directly inhabiting universes.}

We use intrinsically typed syntax \cite{church1940}---terms and types are well-formed by construction.
For conciseness, typing rules omit obvious premises required for well-formedness.
For example, \ruleref{n/tm/lam} implicitly requires that the context be well-typed
($\goodCtx{\Gm}{}$).

%We briefly review the base MLTT fragment first.
%
% In keeping with the recent trend~\cite{altkap2016,coquand2018canonicity,gratzer-sterling-birkedal-2019}\YZ{Does coquand2018canonicity really use explicit subst?}\EDJreply{Yes it is. Look at Section 2.1. When he starts reasoning, he actually deals with category with family (instead of the typing rules), where he consider morphism between contexts, which is substitution. But I agree it can be a bit ambiguous to see (because he literally show two version of MLTT without connecting them---one classical typing rule version, the other directly works on Category with Family (CwF)) but since his proof is mainly on Category with Family, so it counts as explicit substitution. But if you are concerned, we can change to Andras Kovacs "Generalized Universe Hierarchies and First-Class Universe Levels". or Andreas Abel's "Decidability of Conversion for Type Theory in Type Theory". The former one directly work on CwF, the latter one include substitution as part of the raw syntax.}
% in MLTT presentations, we use \emph{explicit substitutions}~\cite{substcalculus,abadi1989subst}:
% substitutions~$\gm$ and their applications (e.g., \sub{T}{\gm}) are part
% of the syntax rather than meta-operations.
%
% Variables are in the form of \citeauthor{debruijn1964} indices:
% \var{n} is the variable bound by the $n$-th closest enclosing binder.
% For example, $\lambda x.\,\lambda y.\,x$ is $\lam{\lam{\var{1}}}$.
%
% Substitutions are typed with the form 
%\EDJ{It seems to me G |- g : D is a judgement not a rule. Rules usually have two parts (premises + conclusion). Judgment is a standard terminology invented by Martin Lof (https://ncatlab.org/nlab/show/judgment see reference here). I will suggest saying 'Formation rules for substitution lead to a judgement G |- g : D' sth like this. In pen-and-paper, the formation rule (corresponding to QIIT well-typed-ness of judgement) is called presuppose (can be found in Modal MLTT paper) }
% $\goodSub{\Gm}{\gm}{\Dl}$.
% The idea is that applying~$\gm$ to terms valid in the context~$\Dl$
% yields terms valid in~$\Gm$ (\ruleref{n/tm/sub} and \ruleref{n/ty/sub}).
% %
% The two main forms of substitutions are weakening (\ruleref{n/sub/wk}) and extension (\ruleref{n/sub/ext}):
% \sub{t}{\SubstWeak{n}} introduces $n$ free variables into the context of $t$, and
% \sub{t}{\SubstExt{\gm}{t'}} substitutes~$t'$ for~$\var{0}$ in~$t$ and
% then applies~$\gm$.
% For example, rule \ruleref{n/tm/snd} states that if $t$ is a dependent pair
% that has type $\TySigma{A}{B}$, then $\snd{t}$ has type
% $\sub{B}{\SubstExt{\SubstWeak{0}}{\fst{t}}}$,
% where $\SubstId$ is the identity substitution (\ruleref{n/sub/id}).

@@

In rule \ruleref{l/add}, the second premise $\goodTerm{\Gm, \mathit{self} : A}{t}{T}$ is
responsible for late binding.
Here, $A$ abstracts the context of the current field~$t$, controlling
how the types of the fields prior to~$t$ are exposed to the typing of~$t$.
Crucially, %unlike the second field in a dependent pair (\ruleref{tm/pair}),
the premise $\goodTerm{\Gm, \mathit{self} : A}{t}{T}$ makes clear that
the typing of~$t$ in \ruleref{l/add} is \emph{universally} quantified---%
rather than \emph{existentially} quantified as is in \ruleref{n/tm/pair}---%
over how the fields in $t$'s context are defined.
Late binding enables reuse.
A different linkage~$\ell'$ that overrides fields in $t$'s context (and thus models a derived family)
can reuse~$t$---without retyping it---by first
projecting~$t$ from $\LkgAdd{\lkg}{t}$ and then appending it to~$\lkg'$.
%\EDJ{I suggest change this sentence to `by projecting it out of parent linkage and adding it in the derived linkage'}.

@@

\input{stlc-linkage-typing.tex}

@@

$$
\Rule[name=Subst]
{\goodType{\Delta}{T}{j} 
  \quad {\goodSub{\Gamma}{\gamma}{\Delta}}}
{\goodType{\Gamma}{T[\gamma]}{j}},
\Rule[]
{\goodTerm{\Delta}{t}{T}
  \quad {\goodSub{\Gamma}{\gamma}{\Delta}}}
{\goodTerm{\Gamma}{t[\gamma]}{T[\gamma]}}
\quad
\Rule[name=Dbj Indices]
{}
{\goodSub{\Gamma, A}{p^1 \gamma}{\Gamma}},
\Rule[]
{}
{\goodSub{\Gamma, A}{"var"_0}{A[p^1]}}
$$

@@

  \quad \goodType{\Gm}{A}{i}
  \quad \goodType{\Gm, A}{B}{i}}

@@

{\goodType{\Gm}{\TyWSingle {\wsig}}{i}

@@

% linkage, signature and compilation and seal
$$
\Rule[name=lsig/ept]
{""}
{\goodSig{\Gm}{\LSigEmp}{0}}
\quad
\Rule[name=ty/lkg]
{\goodSig{\Gm}{\lsig}{n}}
{\goodType{\Gm}{\TyLkg{\lsig}}{i}
\\ \goodType{\Gm}{\TyTkg{\lsig}}{}}
\quad
\Rule[name=lnkg/pkg]
{ \goodTerm{\Gm}{o}{\TyLkg {\lsig}} 
}
{
  \goodTerm{\Gm}{\Tkg {o}}{\TyTkg{\lsig}}
}
$$
$$
\Rule[name=lsig/add]
{\goodSig{\Gm}{\lsig}{n} 
 \quad \goodType{\Gm}{A}{}
 \quad \goodSeal{\Gm}{\seal}{\lsig}{A}
 \quad \goodType{\Gm, A}{T}{i}}
{\goodSig{\Gm}{(\LSigAdd {\lsig}{\seal}{T})}{n+1}}
\quad
\Biggl( \boxed{\goodSeal{\Gamma}{s}{\sigma}{A}} 
:= \boxed{\goodTerm{\Gamma, \cC\sigma}{s}{A[p^1]}} \Biggr)
$$
$$
\Rule[name=lnkg/ept]
{""}
{\goodTerm{\Gm}{\LkgEmp}{\TyLkg {\LSigEmp}}  
}
\quad
\Rule[name=lkg/add]
{ \goodTerm{\Gm}{o}{\TyLkg {\lsig}} 
\\ \goodType{\Gm}{A}{}
\\  \goodSeal{\Gm}{\seal}{\lsig}{A} 
\\  \boxed{\goodTerm{\Gm, A}{t}{T}}
}
{\goodTerm{\Gm}{(\LkgAdd {o} {t})}{\TyLkg{\LSigAdd {\lsig} {\seal} {T}}}
}
$$
$$
\Rule[name=lkg/proj]
{\goodTerm{\Gm}{o}{\TyLkg{\lsig}}}
{\goodTerm{\Gm}{\lkgproj{1} {o}}{\TyLkg {(\lsigproj{1} {\lsig})}}
\\ \boxed{\goodTerm{\Gm, \lsigprojT {\lsig}}{\lkgproj{2}{o}}{\lsigproj{2}{\lsig}}}
}
$$

>>>>

\cref{fig:mltt-selected} presents the syntax and selected typing rules of MLTT %(predicative) MLTT
(and \cref{fig:fmltt-selected} \TT).
Dependent function types \nTyPi{x}{A}{B},
dependent product types \nTySigma{x}{A}{B}, and
identity types \TyId{t_1}{t_2} are standard.
So are their introduction and elimination forms.
Capture-avoiding substitution is notated $\bullet[\bullet/x]$.
A singleton type $\TyS{t}$ helps expose the definition of a term~$t$
in its type (rule \ruleref{tm/s})~\cite{aspinall1995singleton,stone2000}.
For each derivation (of each judgment) we introduce judgmental equality
(of the form \goodTerm{\Gm}{t_1 \equiv t_2}{T})
indicating convertibility.

We use Tarski-style universes \cite{hofmann1997syntax},
which, unlike Russell-style ones,
are not inhabited by types directly but rather by the \emph{codes} of types
and arguably behave better \cite{luo2012notes}.
Term $\codety{T}$ encodes type~$T$ (\ruleref{tm/c}) and thus inhabits universe~$\cU$.
Type $\El{t}$ decodes term~$t$ (\ruleref{ty/el}).
%\footnote{By contrast,  Russel-style universes are inhabited by types directly.  Tarski-style universes are ``semantically more fundamental'' \cite{luo2012notes} and friendlier to our meta-theoretical reasoning.}
There is an infinite hierarchy of universes;
we omit universe levels in the presentation for conciseness.
%\rA{Why is W(t) a term and not a type?}\YZreply{I think the reviewer is wondering what is gained by encoding and decoding types, compared to types directly inhabiting universes.}

% We use intrinsically typed syntax \cite{church1940}---terms and types are well-formed by construction.
For conciseness, typing rules omit obvious premises required for well-formedness.
For example, \ruleref{tm/lam} implicitly requires that the context be well-typed
($\goodCtx{\Gm}{}$).

%We briefly review the base MLTT fragment first.
%
% In keeping with the recent trend~\cite{altkap2016,coquand2018canonicity,gratzer-sterling-birkedal-2019}\YZ{Does coquand2018canonicity really use explicit subst?}\EDJreply{Yes it is. Look at Section 2.1. When he starts reasoning, he actually deals with category with family (instead of the typing rules), where he consider morphism between contexts, which is substitution. But I agree it can be a bit ambiguous to see (because he literally show two version of MLTT without connecting them---one classical typing rule version, the other directly works on Category with Family (CwF)) but since his proof is mainly on Category with Family, so it counts as explicit substitution. But if you are concerned, we can change to Andras Kovacs "Generalized Universe Hierarchies and First-Class Universe Levels". or Andreas Abel's "Decidability of Conversion for Type Theory in Type Theory". The former one directly work on CwF, the latter one include substitution as part of the raw syntax.}
% in MLTT presentations, we use \emph{explicit substitutions}~\cite{substcalculus,abadi1989subst}:
% substitutions~$\gm$ and their applications (e.g., \sub{T}{\gm}) are part
% of the syntax rather than meta-operations.
%
% Variables are in the form of \citeauthor{debruijn1964} indices:
% \var{n} is the variable bound by the $n$-th closest enclosing binder.
% For example, $\lambda x.\,\lambda y.\,x$ is $\lam{\lam{\var{1}}}$.
%
% Substitutions are typed with the form 
%\EDJ{It seems to me G |- g : D is a judgement not a rule. Rules usually have two parts (premises + conclusion). Judgment is a standard terminology invented by Martin Lof (https://ncatlab.org/nlab/show/judgment see reference here). I will suggest saying 'Formation rules for substitution lead to a judgement G |- g : D' sth like this. In pen-and-paper, the formation rule (corresponding to QIIT well-typed-ness of judgement) is called presuppose (can be found in Modal MLTT paper) }
% $\goodSub{\Gm}{\gm}{\Dl}$.
% The idea is that applying~$\gm$ to terms valid in the context~$\Dl$
% yields terms valid in~$\Gm$ (\ruleref{tm/sub} and \ruleref{ty/sub}).
% %
% The two main forms of substitutions are weakening (\ruleref{sub/wk}) and extension (\ruleref{sub/ext}):
% \sub{t}{\SubstWeak{n}} introduces $n$ free variables into the context of $t$, and
% \sub{t}{\SubstExt{\gm}{t'}} substitutes~$t'$ for~$\var{0}$ in~$t$ and
% then applies~$\gm$.
% For example, rule \ruleref{tm/snd} states that if $t$ is a dependent pair
% that has type $\TySigma{A}{B}$, then $\snd{t}$ has type
% $\sub{B}{\SubstExt{\SubstWeak{0}}{\fst{t}}}$,
% where $\SubstId$ is the identity substitution (\ruleref{sub/id}).

@@

In rule \ruleref{l/add}, the second premise $\goodTerm{\Gm, \mathit{self} : A}{t}{T}$ is
responsible for late binding.
Here, $A$ abstracts the context of the current field~$t$, controlling
how the types of the fields prior to~$t$ are exposed to the typing of~$t$.
Crucially, %unlike the second field in a dependent pair (\ruleref{tm/pair}),
the premise $\goodTerm{\Gm, \mathit{self} : A}{t}{T}$ makes clear that
the typing of~$t$ in \ruleref{l/add} is \emph{universally} quantified---%
rather than \emph{existentially} quantified as is in \ruleref{tm/pair}---%
over how the fields in $t$'s context are defined.
Late binding enables reuse.
A different linkage~$\ell'$ that overrides fields in $t$'s context (and thus models a derived family)
can reuse~$t$---without retyping it---by first
projecting~$t$ from $\LkgAdd{\lkg}{t}$ and then appending it to~$\lkg'$.
%\EDJ{I suggest change this sentence to `by projecting it out of parent linkage and adding it in the derived linkage'}.

@@

\input{stlc-linkage-typing-named.tex}

@@

$$
\Rule[name=Subst]
{\goodType{\Delta}{T}{} 
  \quad {\goodSub{\Gamma}{\gamma}{\Delta}}}
{\goodType{\Gamma}{T[\gamma]}{}},
\Rule[]
{\goodTerm{\Delta}{t}{T}
  \quad {\goodSub{\Gamma}{\gamma}{\Delta}}}
{\goodTerm{\Gamma}{t[\gamma]}{T[\gamma]}}
\quad
\Rule[name=Dbj Indices]
{}
{\goodSub{\Gamma, A}{p^1 \gamma}{\Gamma}},
\Rule[]
{}
{\goodSub{\Gamma, A}{"var"_0}{A[p^1]}}
$$

@@

  \quad \goodType{\Gm}{A}{}
  \quad \goodType{\Gm, A}{B}{}}

@@

{\goodType{\Gm}{\TyWSingle {\wsig}}{}

@@

% linkage, signature and compilation and seal
$$
\Rule[name=lsig/ept]
{""}
{\goodSig{\Gm}{\LSigEmp}{0}}
\quad
\Rule[name=ty/lkg]
{\goodSig{\Gm}{\lsig}{n}}
{\goodType{\Gm}{\TyLkg{\lsig}}{}
\\ \goodType{\Gm}{\TyTkg{\lsig}}{}}
\quad
\Rule[name=lnkg/pkg]
{ \goodTerm{\Gm}{o}{\TyLkg {\lsig}} 
}
{
  \goodTerm{\Gm}{\Tkg {o}}{\TyTkg{\lsig}}
}
$$
$$
\Rule[name=lsig/add]
{\goodSig{\Gm}{\lsig}{n} 
 \quad \goodType{\Gm}{A}{}
 \quad \goodSeal{\Gm}{\seal}{\lsig}{A}
 \quad \goodType{\Gm, A}{T}{}}
{\goodSig{\Gm}{(\LSigAdd {\lsig}{\seal}{T})}{n+1}}
\quad
\Biggl( \boxed{\goodSeal{\Gamma}{s}{\sigma}{A}} 
:= \boxed{\goodTerm{\Gamma, \cC\sigma}{s}{A[p^1]}} \Biggr)
$$
$$
\Rule[name=lnkg/ept]
{""}
{\goodTerm{\Gm}{\LkgEmp}{\TyLkg {\LSigEmp}}  
}
\quad
\Rule[name=lkg/add]
{ \goodTerm{\Gm}{o}{\TyLkg {\lsig}} 
\\ \goodType{\Gm}{A}{}
\\  \goodSeal{\Gm}{\seal}{\lsig}{A} 
\\  \boxed{\goodTerm{\Gm, A}{t}{T}}
}
{\goodTerm{\Gm}{(\LkgAdd {o} {t})}{\TyLkg{\LSigAdd {\lsig} {\seal} {T}}}
}
$$
$$
\Rule[name=lkg/proj]
{\goodTerm{\Gm}{o}{\TyLkg{\lsig}}}
{\goodTerm{\Gm}{\lkgproj{1} {o}}{\TyLkg {(\lsigproj{1} {\lsig})}}
\\ \boxed{\goodTerm{\Gm, \lsigprojT {\lsig}}{\lkgproj{2}{o}}{\lsigproj{2}{\lsig}}}
}
$$
